Fail Faster

### 2025-03-21 paper/paper.tex
do a little rewrite of 231-239

Property-based testing (PBT) is a software testing technique that uses
random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often given a very short time
window to find bugs --- e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}

@@

One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
that provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning to find bugs fast.

However, careful measurements show that the performance of existing generator libraries falls significantly short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that the generator uses to make choices can
constitute an unexpectedly high proportion of its runtime.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?

To address this question, we outline two key principles for designing
generator libraries that marry efficiency with flexibility.
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
build DSLs without abstraction overhead (``without regret''~\cn{}).
We apply this technique to generators, completely eliminating the run-time overhead of
many common generator abstractions.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison, showing that showing that certain 
implementations significantly reduce overhead.
Together, these insights highlight practical strategies for improving generator libraries, 
combining the efficiency gains of staging with a careful evaluation of \rand to optimize performance.

@@

The performance impact of this simplification is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
inlined version takes on average half as much time to generate a single pair of
\texttt{int}s. 
Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
show an even more dramatic performance boost.

Because the inlined versions of the generator are identical to the un-inlined
versions except for mechanical, semantics-preserving transformations,
this performance difference is attributable solely to
the different machine code generated by the compiler.
Indeed, compilers of effectful and strict functional languages (including JIT
compilers, in the case of Scala 3)
use heuristics to determine if and when to perform this particular kind of
simplification~\footnote{As we discuss in Section~\ref{section:other-langs}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.}.  Even in cases as simple as
Figure~\ref{fig:simplify}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.

@@

While neither of these microbenchmarks exhibit this behavior, more complex
generators suffer further performance penalties due to \emph{closure
allocation}. In cases where the compiler cannot statically eliminate it,
running a monadic bind allocates a short-lived closure: we allocate a closure
for the continuation, and then immediately jump into it.
In strict functional languages like OCaml and Scala,
Each individual closure allocation is relatively cheap, but doing lots of allocation in a
generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
This effect is also magnified in recursive generators: each iteration
through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
\emph{per generated value} scales linearly with the number of recursive generator calls.

@@

This section presents our experimental setup, as well as answers to the following research questions:

@@

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a well-typed generator of
STLC terms, whose behavior is the most complex. 
For evaluating \scalaname, we implement a subset of the aformentioned 
strategies---the boolean list generator, single-pass BST generator, and 
well-typed STLC term generator.

>>>>

Property-based testing (PBT) is a software testing technique which uses
random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often given a very short time
window to find bugs --- e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
In other words, the goal is to fail as fast as possible. \jwc{First cut at ``fail fast''}

@@

Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
that provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning to find bugs fast.

However, careful measurements show that the performance of existing generator libraries falls significantly short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design that gives generator libraries their
flexibility and expressiveness also introduces overhead: layered abstractions
and indirections hinder compiler optimizations, shifting work to
run time.
Second, the \rand that underlies the generator's operation---the source of the
random numbers that are used to make choices in the generator---can
constitute an unexpectedly high proportion of a generator's running time.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?

To address this question, we outline two key principles for designing
generator libraries that balance efficiency with flexibility.\bcp{We
  just said we want both, not a ``balance'' that trades one off
  against the other...}
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
build DSLs without abstraction overhead (``without regret''~\cn{}).
We apply this technique to generators, completely eliminating the run-time overhead of
many common generator abstractions.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison.
Together, we refer to this recipe\bcp{``demonstrating the impact'' is
  not a recipe} for improving generator libraries as \name{} --- \jwc{something pithy here}.

@@

The performance impact of this simplification is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
inlined version takes on average half as much time to generate a single pair of
\texttt{int}s. 
Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
show an even more dramatic performance boost.
Because the inlined versions of the generator are identical to the un-inlined
versions except for mechanical, semantics-preserving transformations,
this performance difference is attributable solely to
the different machine code generated by the compiler (and not lucky choice of random seeds).

It might be surprising to readers that this dramatic overhead exists --- shouldn't the compiler perform
this simple optimization? Compilers of effectful and strict functional languages (including JIT
compilers in the case of Scala 3)
do in fact have heuristics to determine if and when to perform this particular kind of
simplification\footnote{As we discuss in Section~\ref{section:other-langs}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.}. However even in cases as simple as
Figure~\ref{fig:simplify}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.
\jwc{Rewrite this para, BCP what do you think now?}

@@

More complex
generators suffer further performance penalties due to \emph{closure
allocation}. In cases where the compiler cannot statically eliminate it,
running a monadic bind allocates a short-lived closure
for the continuation, and then immediately jumps into it.
In strict functional languages like OCaml and Scala,
Each individual closure allocation is relatively cheap, but doing lots of allocation in a
generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
This effect is also magnified in recursive generators: each iteration
through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
\emph{per generated value} scales linearly with the number of recursive generator calls.

@@

This section presents our experimental setup, and answers to the following research questions:

@@

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a well-typed generator of
STLC terms, whose behavior is the most complex. 
For evaluating \scalaname, we implement a subset of the aformentioned 
strategies---only the boolean list generator, single-pass BST generator, and 
well-typed STLC term generator.
